Visualizing the Geography of Genetic Variants

### 2016-07-01 paper.Rmd
reformatted figures bib and fixed some typos

The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry. Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography. Thus by surveying the collection of rare variants that an individual carries provides information about the population origins of that individual's recent ancestors. One of us (J.N.) recently took advantage of the geographic clustering of an individual's rare variants in helping to reaffirm the historically highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.

Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets. For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide vairants (SNVs). One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast. While these approaches are extremely valuable, researchers can use them to often without inspecting the underlying variant data in more detail. A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.

@@

Our work is inspired by past tools such as the ALFRED database @rajeevan2011alfred and the maps available on the HGDP Selection browser @pickrell2009signals.  One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a sbustantial learning curve.  The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @pickrell2009signals, @coop2009role, books e.g. @dudley2013exploring, and have been ported and made avialable on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).

@@

# Underlying Frequency API {-}
To interface population minor allele frequency data stored in the SQL database with the front-end D3.js visualization we use a REST API impelemented in the python library Flask-RESTful @citeXXX. GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g. latitude, longitude, population label, sample size, and frequency scale. Genetic variants can be queried by chromosome position, rsid, or randomly. Example HTTP requests and json response can be seen below:

@@

# Caveats {-}
A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.  While doing so is generally immensely helpful, it has inherent complexity.  For example,  pracitioners must make choices regarding representing where an individual was sampled for the study (e.g. the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g. as determined in practice by the birthplaces of recent ancestors, such as grandparents).  We do not proscribe a general solution to this problem and instead use a set of locations for each dataset that aligns with those used by the initial analysts of the data.  A future feature will be to allow alternative location schema to be used for the populations in a dataset.

@@

>>>>

The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry. Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography. By surveying the collection of rare variants that an individual carries one can gain insight into the geographic origins of that individual's recent ancestors. One of us (J.N.) recently took advantage of the geographic localization of an individual's rare variants in helping to reaffirm the historically-highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.

Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets. For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs). One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast. While these approaches are extremely valuable, researchers can use them to often without inspecting the underlying variant data in more detail. A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.

@@

Our work is inspired by past tools such as the ALFRED database @rajeevan2011alfred and the maps available on the HGDP Selection browser @pickrell2009signals.  One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a substantial learning curve.  The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @pickrell2009signals, @coop2009role, books e.g. @dudley2013exploring, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).

@@

# Underlying Frequency API {-}
To interface population minor allele frequency data stored in the SQL database with the front-end D3.js visualization we use a REST API implemented in the python library Flask-RESTful @citeXXX. GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g. latitude, longitude, population label, sample size, and frequency scale. Genetic variants can be queried by chromosome position, rsid, or randomly. Example HTTP requests and json response can be seen below:

@@

# Caveats {-}
A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.  While doing so is generally immensely helpful, it has inherent complexity.  For example,  practitioners must make choices regarding representing where an individual was sampled for the study (e.g. the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g. as determined in practice by the birthplaces of recent ancestors, such as grandparents).  We do not proscribe a general solution to this problem and instead use a set of locations for each dataset that aligns with those used by the initial analysts of the data.  A future feature will be to allow alternative location schema to be used for the populations in a dataset.

@@

<!-- Has to be set for some wierd indentation issue -->
\setlength\parindent{0pt}

### 2016-07-01 paper.tex
reformatted figures bib and fixed some typos

The geographic distribution of rare genetic variants (minor allele
frequency \(\leq .05\)) can be particularly informative in determining
an individual's ancestry. Rare variants tend to be localized in small
geographic regions because they have arisen recently in human history
and have had little time to spread across populations through migration
events {[}4{]}, {[}5{]}. Thus by surveying the collection of rare
variants that an individual carries provides information about the
population origins of that individual's recent ancestors. One of us
(J.N.) recently took advantage of the geographic clustering of an
individual's rare variants in helping to reaffirm the historically
highly contested Native American ancestry of the Kennewick Man, an
ancient individual from Washington State \textbf{???}.

Contemporary population genetics researchers are also faced with the
challenge of large, high dimensional datasets. For example, is not
uncommon for a researcher in human genetics to have a dataset comprised
of thousands of individuals measured at hundreds of thousands or even
millions of single nucleotide vairants (SNVs). One common approach to
visualizing such high-dimensional data is to compress the SNV dimensions
down to a small number of latent factors, using a method such as
principal components analysis {[}6{]}, {[}7{]}, a model-based clustering
method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While these
approaches are extremely valuable, researchers can use them to often
without inspecting the underlying variant data in more detail. A natural
approach to gaining more insight to the overall structure of a
population genetic dataset is to visually inspect what geographic
patterns arise in allele frequency maps.

@@

Our work is inspired by past tools such as the ALFRED database {[}11{]}
and the maps available on the HGDP Selection browser {[}12{]}. One of us
(J.N.) developed the scripts for the HGDP Selection Browser maps using
The Generic Mapping Tools (GMT) {[}13{]}, a powerful system of
geographic plotting scripts with a sbustantial learning curve. The plots
from the HGDP Selection Browser have proved useful, and have appeared in
research articles e.g {[}12{]}, {[}14{]}, books e.g. {[}15{]}, and have
been ported and made avialable on the UCSC Genome Browser (available
under the HGDP Allele Freq track of the browser).

@@

To interface population minor allele frequency data stored in the SQL
database with the front-end D3.js visualization we use a REST API
impelemented in the python library Flask-RESTful \textbf{???}. GET
requests are submitted to the database via HTTP, returning json
formatted allele frequency data and the meta-data associated with each
population and genetic variant e.g.~latitude, longitude, population
label, sample size, and frequency scale. Genetic variants can be queried
by chromosome position, rsid, or randomly. Example HTTP requests and
json response can be seen below:

@@

A major challenge of using a geographic representation of genetic
variation in humans is that the samples must be associated with a
geographic location. While doing so is generally immensely helpful, it
has inherent complexity. For example, pracitioners must make choices
regarding representing where an individual was sampled for the study
(e.g.~the city of a major research center) or choosing a location that
is more representative of an individual's ancestral origins (e.g.~as
determined in practice by the birthplaces of recent ancestors, such as
grandparents). We do not proscribe a general solution to this problem
and instead use a set of locations for each dataset that aligns with
those used by the initial analysts of the data. A future feature will be
to allow alternative location schema to be used for the populations in a
dataset.

@@

>>>>

The geographic distribution of rare genetic variants (minor allele
frequency \(\leq .05\)) can be particularly informative in determining
an individual's ancestry. Rare variants tend to be localized in small
geographic regions because they have arisen recently in human history
and have had little time to spread across populations through migration
events {[}4{]}, {[}5{]}. By surveying the collection of rare variants
that an individual carries one can gain insight into the geographic
origins of that individual's recent ancestors. One of us (J.N.) recently
took advantage of the geographic localization of an individual's rare
variants in helping to reaffirm the historically-highly contested Native
American ancestry of the Kennewick Man, an ancient individual from
Washington State \textbf{???}.

Contemporary population genetics researchers are also faced with the
challenge of large, high dimensional datasets. For example, is not
uncommon for a researcher in human genetics to have a dataset comprised
of thousands of individuals measured at hundreds of thousands or even
millions of single nucleotide variants (SNVs). One common approach to
visualizing such high-dimensional data is to compress the SNV dimensions
down to a small number of latent factors, using a method such as
principal components analysis {[}6{]}, {[}7{]}, a model-based clustering
method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While these
approaches are extremely valuable, researchers can use them to often
without inspecting the underlying variant data in more detail. A natural
approach to gaining more insight to the overall structure of a
population genetic dataset is to visually inspect what geographic
patterns arise in allele frequency maps.

@@

Our work is inspired by past tools such as the ALFRED database {[}11{]}
and the maps available on the HGDP Selection browser {[}12{]}. One of us
(J.N.) developed the scripts for the HGDP Selection Browser maps using
The Generic Mapping Tools (GMT) {[}13{]}, a powerful system of
geographic plotting scripts with a substantial learning curve. The plots
from the HGDP Selection Browser have proved useful, and have appeared in
research articles e.g {[}12{]}, {[}14{]}, books e.g. {[}15{]}, and have
been ported and made available on the UCSC Genome Browser (available
under the HGDP Allele Freq track of the browser).

@@

To interface population minor allele frequency data stored in the SQL
database with the front-end D3.js visualization we use a REST API
implemented in the python library Flask-RESTful \textbf{???}. GET
requests are submitted to the database via HTTP, returning json
formatted allele frequency data and the meta-data associated with each
population and genetic variant e.g.~latitude, longitude, population
label, sample size, and frequency scale. Genetic variants can be queried
by chromosome position, rsid, or randomly. Example HTTP requests and
json response can be seen below:

@@

A major challenge of using a geographic representation of genetic
variation in humans is that the samples must be associated with a
geographic location. While doing so is generally immensely helpful, it
has inherent complexity. For example, practitioners must make choices
regarding representing where an individual was sampled for the study
(e.g.~the city of a major research center) or choosing a location that
is more representative of an individual's ancestral origins (e.g.~as
determined in practice by the birthplaces of recent ancestors, such as
grandparents). We do not proscribe a general solution to this problem
and instead use a set of locations for each dataset that aligns with
those used by the initial analysts of the data. A future feature will be
to allow alternative location schema to be used for the populations in a
dataset.

@@

\setlength\parindent{0pt}
